File-based localization of numerical perturbations in data analysis pipelines

### 2020-03-23 paper.tex
Updates and comments in tool description

% There are different approaches to improve the reproducibility of the analysis, 
% including containerization techniques that encapsulate software/hardware dependencies, 
% provenance capturing tools, and version control systems. 
% However, a comprehensive solution requires to fix the numerical instabilities instead of 
% masking the problem. For this purpose, bagging is one solution that has been used to 
% stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
% Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
% As the second solution, a debugging tool can help to identify and 
% fix the instabilities in the pipeline. Therefore, we introduce 
% the numerical reproducibility measurement tool named \emph{Spot} to identify 
% the processes in the pipeline that create numerical differences across different conditions. 
% Furthermore, we can stabilize the identified processes in the next steps.

@@

The steps implemented in \toolname to compare pipeline executions in
2 conditions are outlined in Figure~\ref{fig:overview-tool}. First,
in Condition 1, a directed, rooted, bipartite provenance graph is built for
each subject, where nodes represent application processes or files, and
edges represent read or write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). \emph{Transient} files are then identified as
 files that are either deleted during the execution, or modified by
multiple processes (Step 2). \toolname re-runs the pipeline to save these files
right after they are produced, resulting in a provenance \emph{DAG}
(Directed Acyclic Graph) in which file nodes have a single parent (in-degree
of 1) (Step 3). This DAG is then compared to the one produced from
Condition 2, in a step-by-step process that prevents the propagation of differences (step 4).
\toolname finally produces a labeled graph that identifies the
non-reproducible processes of the pipeline.

\begin{figure*}
  \centering
    \includegraphics[width=.8\textwidth]{images/spot-diagram}
    \caption{An overview of Spot tool in four steps.
    (\href{https://docs.google.com/drawings/d/1OB3sB8kkK17Q516-TcmXKh3TwXSLDToJB11IVfUH5AU/edit?usp=sharing}{Source})}
    \label{fig:overview-tool}
  \end{figure*}

  \todo{review and summarize assumptions:
  * No concurrent writes from multiple processes
  * A given process cannot modify the bytes it wrote to a file
  * Process are command lines
  But we support non-concurrent writes by different processes. This is common in neuroimaging.
  }

@@

We use the \reprozip tool~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
\toolname then reconstructs a provenance graph by creating process and file
nodes from the \reprozip trace, and by adding directed edges corresponding
to file reads and file writes (Figure~\ref{fig:simple_script}-left).

\todo{Is every process linked to an executable?}

@@

\begin{figure}
\centering
  \includegraphics[width=1\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  constructed from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Nodes are devided into two sets including processes that are shaped 
  in the circles and files that are shaped in the boxes.
  Transient files are represented with grey color boxes. 
  Plain edges 
  represent the process tree (\texttt{fork()} or \texttt{clone()} 
  system calls). Dashed edges represent file dependencies.
  Every node in the graph is labeled using (1) a process id created by our
  reconstruction, (2) the name of the executable run by the process or the file name.
  Process 0 is the initial call to the \texttt{sh} interpreter, and
  the rest of processes are the calls to FSL robustfov, bet, stats and maths made in
  Listing~\ref{listing:sample-script}. Processes 75 and 149 are forked by processes 4 and 78, respectively: 
  they were captured by \reprozip while they did not appear in
  Listing~\ref{listing:sample-script}.  
}
  \label{fig:simple_script}
\end{figure}

@@

\toolname captures temporary files in Condition 1 by replacing every
process P by a wrapper that first calls P and then saves the produced
temporary files to a read-only directory. This process replacement is done
without pipeline instrumentation, by pre-pending to the \texttt{PATH} environment
variable a directory containing a wrapper script named as the executable
called by P.

Files written by multiple processes are disambiguated as follows. For a
 file F written by the processes in \textbf{P} = \{$P_{1}$, \ldots,
 $P_{n}$\}, \toolname first checks that processes in \textbf{P} do not
 write concurrently to F, which would violate our assumptions. Then, it
 replaces every process $P_{i}$ by a \texttt{PATH}-based wrapper that first
 calls $P_{i}$ and then saves F to a read-only directory. In this way,
 successive versions of F are preserved for comparison. \toolname finally
 updates the provenance graph accordingly, so that all files in the graph
 have an in-degree of 1 (Figure~\ref{fig:simple_script}-right). This operation also makes the provenance graph
 acyclic, since we assumed that a process can only write a single time to a file.

As an example, in the sample pipeline in
Listing~\ref{listing:sample-script}, process \texttt{3\#robustfov} writes a
temporary file. Also, file \emph{T1w\_brain.nii.gz} is first written
by \texttt{75\#bet2} and then overwritten by process \texttt{149\#bet2}. 
The complete list of the processes that
write transient files is captured in Listing~\ref{listing:json_process}.
Using this list, we modify processes in Condition 1 and then re-execute
pipeline to save them in a backup folder.

\begin{listing}
  \begin{minted}[frame=single,
    framesep=3mm,
    linenos=false,
    xleftmargin=0pt,
    tabsize=4]{JSON}
{
  "multi_write_process": {
    "bet2\00T1w\00T1w_brain":
    {
      "files":[
        "/cond1/subj1/T1w_brain.nii.gz"
      ],
      "pid": 75
    },
    "bet2\00T1w_roi\00T1w_brain":
    {
      "files":[
        "/cond1/subj1/T1w_brain.nii.gz"
      ],
      "pid": 149
    }
  },
  "temp_process": {
    "robustfov\00-i\00T1w.nii.gz\00-r\00T1w_roi.nii.gz": 
    {
      "files": [
        "/cond1/subj1/T1w_roi.nii.gz"
      ],
      "pid": 3
    } 
  } 
}

\end{minted}
\caption{List of processes and transient files from the example pipeline in 
Listing~\ref{listing:sample-script}.}
\label{listing:json_process}
\end{listing}

@@

Once all the files have been captured in Condition 1, including transient
ones, \toolname labels pipeline processes through a step-by-step execution
in Condition 2. Each process in Condition 2 is run and its output files are
compared to the ones produced in Condition 1. If no differences are found,
the process is marked as reproducible. Otherwise, the process is marked as
non-reproducible and the output files produced in Condition 1 are copied to
Condition 2, to ensure that differences do not propagate further in the
pipeline. Processes are instrumented transparently, through a modification
of the \texttt{PATH} variable similar to the one described previously. By
default, differences in output files are identified by comparing file
checksums. Other comparison functions can also be defined for specific file
types, for instance to ignore sections of the file containing timestamps or
other execution-specific information. Finally, \toolname creates a labeled
provenance graph, highlighting non-reproducible processes.

@@

>>>>

% There are different approaches to improve the reproducibility of the analysis, 
% including containerization techniques that encapsulate software/hardware dependencies, 
% provenance capturing tools, and version control systems. 
% However, a comprehensive solution requires to fix the numerical instabilities instead of 
% masking the problem. For this purpose, bagging is one solution that has been used to 
% stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
% Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
% As the second solution, a debugging tool can help to identify and 
% fix the instabilities in the pipeline. Therefore, we introduce 
% the numerical reproducibility measurement tool named \emph{Spot} to identify 
% the processes in the pipeline that create numerical differences across different conditions. 
% Furthermore, we can stabilize the identified processes in the next steps.
\todo{explain creation vs propagation}
\todo{define pipeline}

@@

\toolname identifies the processes in a pipeline that produce different results in two different conditions.  
In Condition 1, a directed, rooted, bipartite provenance graph is recorded for
each subject, where nodes represent application processes and files, and
edges represent read and write file accesses (Step 1 in Figure~\ref{fig:overview-tool}).
\emph{Transient} files are then identified as
 files that are either deleted during the execution, or modified by
multiple processes (Step 2). \toolname re-runs the pipeline to save these files
right after they are produced, resulting in a provenance \emph{DAG}
(Directed Acyclic Graph) in which file nodes have a single parent (in-degree
of 1) (Step 3). This DAG is then compared to the one produced from
Condition 2, in a step-by-step execution to prevent the propagation of differences (step 4).
\toolname finally produces a labeled graph that identifies the
non-reproducible processes of the pipeline.

\toolname makes the following assumptions on the pipeline, meant to ensure that 
a file can be unambiguously associated with the process that created it:
\begin{enumerate}
  \item Files are not written concurrently by different processes, although sequential writes by multiple processes are supported;
  \item In case a process writes multiple versions of the same file, only a single version is made available to other processes.
\end{enumerate} 
In addition, \toolname also requires that processes are associated to a
command line (executable and arguments), to facilitate process instrumentation.

\tristan{Ali: please review these assumptions and check if any other one is required}

\begin{figure*}
  \centering
    \includegraphics[width=.8\textwidth]{images/spot-diagram}
    \caption{\toolname overview.}
    \label{fig:overview-tool}
  \end{figure*}

@@

\toolname uses the \reprozip tool~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
\toolname then reconstructs a provenance graph by creating process and file
nodes from the \reprozip trace, and by adding directed edges corresponding
to file reads and file writes (Figure~\ref{fig:simple_script}-left). \tristan{the new example is 
confusing: it has too many processes and its purpose is confusing. This should be simplified. 
See suggestion in example.sh: it has only a few processes, and it has a temporary and a multiple-write file. 
If you agree that this example is enough, could you update the figure?}

@@

\begin{figure}
\centering
  \includegraphics[width=1\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  constructed from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles and files with rectangles.
  Transient files are shaded in gray. 
  Plain edges 
  represent the process tree (\texttt{fork()} or \texttt{clone()} 
  system calls) \tristan{Does spot make any use of the process tree at any stage? If not, I think we may want to remove these edges.}. Dashed edges represent file dependencies.
  Every node in the graph is labeled using (1) a process id created by our
  reconstruction, (2) the name of the executable run by the process or the file name.
  Process 0 is the initial call to the \texttt{sh} interpreter, and
  the rest of processes are the calls to FSL robustfov, bet, stats and maths made in
  Listing~\ref{listing:sample-script}. Processes 75 and 149 are forked by processes 4 and 78, respectively: 
  they were captured by \reprozip while they did not appear in
  Listing~\ref{listing:sample-script}.  
}
  \label{fig:simple_script}
\end{figure}

@@

\toolname captures temporary files in Condition 1 by replacing every
process P by a wrapper that first calls P and then saves the produced
temporary files to a read-only directory. This process replacement does not require 
modifying the pipeline: it only pre-pends to the \texttt{PATH} environment
variable a directory containing a wrapper script named after the executable
called by P.

Files written by multiple processes are disambiguated as follows. For a
 file F written by the processes in \textbf{P} = \{$P_{1}$, \ldots,
 $P_{n}$\}, \toolname first checks that processes in \textbf{P} do not
 write concurrently to F, which would violate our assumptions. Then, it
 replaces every process $P_{i}$ by a \texttt{PATH}-based wrapper that first
 calls $P_{i}$ and then saves F to a read-only directory. In this way,
 successive versions of F are preserved for comparison. \toolname finally
 updates the provenance graph accordingly, so that all files in the graph
 have an in-degree of 1 (Figure~\ref{fig:simple_script}-right). This operation also makes the provenance graph
 acyclic, since we assumed that a process could only release a single version of a file.

As an example, in the sample pipeline in
Listing~\ref{listing:sample-script}, process \texttt{3\#robustfov} writes a
temporary file. Also, file \emph{T1w\_brain.nii.gz} is first written
by \texttt{75\#bet2} and then overwritten by process \texttt{149\#bet2}.

@@

Once transient files are captured, \toolname labels pipeline processes
through a step-by-step execution in Condition 2. Each process in Condition
2 is run and its output files are compared to the ones produced in
Condition 1. If no differences are found, the process is marked as
reproducible. Otherwise, the process is marked as non-reproducible and the
output files produced in Condition 1 are copied to Condition 2, to ensure
that differences do not propagate further in the pipeline. Processes are
instrumented transparently, through a modification of the \texttt{PATH}
variable similar to the one described previously. By default, differences
in output files are identified by comparing file checksums. Other
comparison functions can also be defined for specific file types, for
instance to ignore sections of the file containing timestamps or other
execution-specific information. Finally, \toolname creates a labeled
provenance graph, highlighting non-reproducible processes.

@@

* granularity is that of processes, no in-memory checks.
